---
title: "SAMHSA NSSATS modeling and formal testing "
output: html_notebook
---


Checking assumptions of linear regression: normal distribution and continuous variable
```{r}
#all of the forprofit distributions are skewed right 
hist(state_final_df[state_final_df$profit_type=="Forprofit", ]$count_tx)

#all of the nonprofit distributions are skewed right
hist(state_final_df[state_final_df$profit_type=="Nonprofit", ]$count_tx)

#all of the government distributions are skewed right
hist(state_final_df[state_final_df$profit_type=="Government", ]$count_tx)

```


For count data, we have options of poisson or negative binomial distributions. 

Negative binomial distribution is appropriate if there is sparse data (not applicable) or overdispersion (variance > mean)
```{r}
print("For Profit Overdispersion")
var(profit_states_1419$count_Forprofit_2015) > mean(profit_states_1419$count_Forprofit_2015)
var(profit_states_1419$count_Forprofit_2016) > mean(profit_states_1419$count_Forprofit_2016)
var(profit_states_1419$count_Forprofit_2017) > mean(profit_states_1419$count_Forprofit_2017)
var(profit_states_1419$count_Forprofit_2018) > mean(profit_states_1419$count_Forprofit_2018)
var(profit_states_1419$count_Forprofit_2019) > mean(profit_states_1419$count_Forprofit_2019)

print("Non Profit Overdispersion")
var(profit_states_1419$count_Nonprofit_2015) > mean(profit_states_1419$count_Nonprofit_2015)
var(profit_states_1419$count_Nonprofit_2016) > mean(profit_states_1419$count_Nonprofit_2016)
var(profit_states_1419$count_Nonprofit_2017) > mean(profit_states_1419$count_Nonprofit_2017)
var(profit_states_1419$count_Nonprofit_2018) > mean(profit_states_1419$count_Nonprofit_2018)
var(profit_states_1419$count_Nonprofit_2019) > mean(profit_states_1419$count_Nonprofit_2019)

print("Government Overdispersion")
var(profit_states_1419$count_Government_2015) > mean(profit_states_1419$count_Government_2015)
var(profit_states_1419$count_Government_2016) > mean(profit_states_1419$count_Government_2016)
var(profit_states_1419$count_Government_2017, na.rm = TRUE) > mean(profit_states_1419$count_Government_2017, na.rm = TRUE) #there is 1 null value in this data 
var(profit_states_1419$count_Government_2018) > mean(profit_states_1419$count_Government_2018)
var(profit_states_1419$count_Government_2019) > mean(profit_states_1419$count_Government_2019)


```


Test for collinearity among control variables:

needing_tx is correlated with other substance use variables:
  sud_pastyear and needing_tx are HIGHLY correlated
  drug_pastmonth and needing_tx are decently correlated (0.7)
  bingealcohol_pastmonth and needing_tx are decently correlated (0.64)
  
bingealcohol_pastmonth and sud_pastyear are somewhat correlated 


I think it makes sense to only use sud_pastyear for the demand measure, since having the diagnosis may be a better indicator of need anyway (people who use may not necessarily want treatment)
```{r}
cor(state_final_df[, c("pop_thousands", "percapita_health_spending", "medicaid_enroll_percapita", "needing_tx", "sud", "binge_drinking", "illicit_druguse")])
```

https://cran.r-project.org/web/packages/PanelCount/vignettes/vignette.html


Using pglm() to fit a poisson model (neg.bin does not converge)
```{r}
library(lme4)
library(pglm)
library(plm)
library(MASS)


state_forprofit <- state_final_df %>% 
                    arrange(State_Abbrev, year) %>%  # Ensure data is ordered by State_Abbrev and year
                    group_by(State_Abbrev) %>%
                    mutate(prev2years_bingedrinking = dplyr::lag(binge_drinking),
                           prev2years_druguse = dplyr::lag(illicit_druguse))%>%
                    filter(profit_type == "Forprofit", !is.na(prev2years_bingedrinking))
                    


#the previous 2 years' data does significantly predict the number of for-profit treatment centers
poissonmodel_forprofit <- pglm(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending  + 
                                 prev2years_bingedrinking + prev2years_druguse ,
                              #when controlling for all of the demand variables, nothing is significant. I think this is because sud and needing_tx are correlated with other variables (see above)
                              # +lag(sud)  + lag(needing_tx), 
    #the effect should be twooways because there may be state-level differences and country-level differences in each year                      
    effect = "twoways", 
   index = c('State_Name', 'year'),
    data = state_forprofit, 
    model = "within", 
   #model = 'within' means fixed effects, or variable intercept by state by constant slope
    #method = 'nr',
    family = poisson) #can consider using 'negbin' instead

summary(poissonmodel_forprofit)
exp(poissonmodel_forprofit$estimate)

#trying negative binomial - i dont think controlling for the states is the same as fixed effects
#nbmodel_forprofit <- glm.nb(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending  + 
#                                 lag(binge_drinking) + lag(illicit_druguse) + factor(State_Name),
#                             data = state_forprofit)
#summary(nbmodel_forprofit)

# nbmodel_forprofit <- pglm(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending  + 
#                                  prev2years_bingedrinking + prev2years_druguse + factor(year),
#                               #when controlling for all of the demand variables, nothing is significant. I think this is because sud and needing_tx are correlated with other variables (see above)
#                               # +lag(sud)  + lag(needing_tx), 
#     #the effect should be twooways because there may be state-level differences and country-level differences in each year                      
#    index = c('State_Name', 'year'),
#     data = state_forprofit, 
#    model = "within", 
#    print.level=3,
#    #model = 'within' means fixed effects, or variable intercept by state by constant slope
#     method = 'nr',
#     family = negbin)


#try a linear model
lm_forprofit <- plm(tx_percapita ~ medicaid_enroll_percapita + percapita_health_spending  + 
                                 prev2years_bingedrinking + prev2years_druguse ,
                              
   index = c('State_Name', 'year'),
    data = state_forprofit, 
    model = "within") 

summary(lm_forprofit)
lm_forprofit$coefficients


#because the NSDUH data is combined over 2 years, it may make sense to try the CDC overdose deaths data, too
#the previous 1 year's overdose data does significantly predict the number of for-profit treatment centers
# simplemodel_forprofit2 <- pglm(count_tx ~ pop_thousands+ medicaid_enroll_percapita + percapita_health_spending  + sud, 
#     effect = "individual",
#    index = c('State_Name', 'year'),
#     data = state_forprofit, 
#     family = poisson)
# 
# summary(simplemodel_forprofit2)
# exp(simplemodel_forprofit2$estimate)
```

```{r}
#binge drinking and drug use don't seem to be too correlated but seem to become insignificant when controlling for both
#cor(state_forprofit$prev2years_bingedrinking, state_forprofit$prev2years_druguse)
```


Trying to use GLMMadaptive instead to fit a negative binomial model with country-level grouping
https://drizopoulos.github.io/GLMMadaptive/


Getting an error of "the leading minor of order 1 is not positive" when i introduce more covariates 

```{r}
library(GLMMadaptive)

#example from website
#https://drizopoulos.github.io/GLMMadaptive/reference/negative_binomial.html
#set.seed(102)
#dd <- expand.grid(f1 = factor(1:3), f2 = LETTERS[1:2], g = 1:30, rep = 1:15,
#                  KEEP.OUT.ATTRS = FALSE)
#mu <- 5 * (-4 + with(dd, as.integer(f1) + 4 * as.numeric(f2)))
#dd$y <- rnbinom(nrow(dd), mu = mu, size = 0.5)#

#gm1 <-  mixed_model(fixed = y ~ f1 * f2, random = ~ 1 | g, data = dd, 
#                    family = GLMMadaptive::negative.binomial())

#summary(gm1)

fm <- mixed_model(fixed = count_tx ~ prev2years_druguse + prev2years_bingedrinking,
                     #percapita_health_spending, #medicaid_enroll_percapita + ,
                    #pop_thousands +   + 
                                 #
                  random = ~ 1 | State_Name, 
                  data = state_forprofit,
                  family = GLMMadaptive::negative.binomial(),
                  #initial_values = list(betas = poisson()), #rep(0, 3)),
                  #iter_EM = 0,
                  #penalized = TRUE,
                  max_coef_value= 10000
                  )

summary(fm)
```

Let's also try glmer model?

Because the percapita_health_spending and pop_thousands are so large relative to the prev2years_drug use and medicaid_enroll_percapita, the model struggles with convergence. we need to scale these variables. 
```{r}
library(lme4)
#scale to health spending in thousands of dollars
state_forprofit$percapita_health_spending_thousands <- state_forprofit$percapita_health_spending / 1000
#scale to population in millions 
state_forprofit$pop_millions <- state_forprofit$pop_thousands / 1000


#this model throws a warning about the predictors being on different scales 
glmer_forprofit1 <- glmer.nb(count_tx ~ prev2years_druguse + #prev2years_bingedrinking + #primary predictor of interest
           medicaid_enroll_percapita + 
           percapita_health_spending_thousands + 
           pop_millions +
           (1|State_Name)  #random effects 
           ,
          #nAGQ=0,
         data = state_forprofit)
summary(glmer_forprofit1)


#this model does not throw a warning, and displays a lower BIC
glmer_forprofit2 <- glmer.nb(count_tx ~ prev2years_druguse + #prev2years_bingedrinking + #primary predictor of interest
          medicaid_enroll_percapita +
           log(percapita_health_spending) + #scale using log
           log(pop_thousands) + #scale using log
           (1|State_Name)  # random effects 
           ,
          #nAGQ=0,
         data = state_forprofit)
summary(glmer_forprofit2)

#when we use this model and control for the year, we see that most of the variation is explained by the year alone
#however, this does not converge
glmer_forprofit3 <- glmer.nb(count_tx ~ prev2years_druguse + #prev2years_bingedrinking + #primary predictor of interest
           I(as.numeric(year) - 2017) +
           medicaid_enroll_percapita +
           log(percapita_health_spending) + #scale using log
           log(pop_thousands)+ #scale using log
           (1|State_Name)  # random effects 
           ,
          #nAGQ=0,
         data = state_forprofit)
summary(glmer_forprofit3)

```


#may want to consider using glmmTMB later on, AND VALIDATE RESULTS THIS WAY  per https://stackoverflow.com/questions/65565135/model-convergence-warning-with-negative-binomial-glmer





```{r}
anova(glmer_forprofit1, glmer_forprofit2, glmer_forprofit3)
```


```{r}
state_nonprofit <- state_final_df %>% 
                    arrange(State_Abbrev, year) %>%  # Ensure data is ordered by State_Abbrev and year
                    group_by(State_Abbrev) %>%
                    mutate(prev2years_bingedrinking = dplyr::lag(binge_drinking),
                           prev2years_druguse = dplyr::lag(illicit_druguse))%>%
                    filter(profit_type == "Nonprofit", !is.na(prev2years_bingedrinking))


poissonmodel_nonprofit <- pglm(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending  + 
                                 prev2years_bingedrinking + prev2years_druguse ,
                              #when controlling for all of the demand variables, nothing is significant. I think this is because sud and needing_tx are correlated with other variables (see above)
                              # +lag(sud)  + lag(needing_tx), 
    #the effect should be twooways because there may be state-level differences and country-level differences in each year                      
    effect = "twoways", 
   index = c('State_Name', 'year'),
    data = state_nonprofit, 
    model = "within", 
   #model = 'within' means fixed effects, or variable intercept by state by constant slope
    #method = 'nr',
    family = poisson) #can consider using 'negbin' instead

summary(poissonmodel_nonprofit)


#we could also try fitting this as a linear regression with panel data? to predict treatment centers per capita
linear_model <- plm(tx_percapita ~ medicaid_enroll_percapita + percapita_health_spending + lag(sud_pastyear),
    index = c('State_Name', 'year'),
    effect = c("twoways"), #we want effects for both individual states and time varying
    model = "within",
    data = state_nonprofit)

summary(linear_model)

linear_model2 <- plm(tx_percapita ~ medicaid_enroll_percapita + percapita_health_spending + lag(RATE),
    index = c('State_Name', 'year'),
    effect = c("twoways"), #we want effects for both individual states and time varying
    model = "within",
    data = state_nonprofit)

summary(linear_model2)
```


```{r}

state_govt <- state_final_df %>% 
                    arrange(State_Abbrev, year) %>%  # Ensure data is ordered by State_Abbrev and year
                    group_by(State_Abbrev) %>%
                    mutate(prev2years_bingedrinking = dplyr::lag(binge_drinking),
                           prev2years_druguse = dplyr::lag(illicit_druguse))%>%
                    filter(profit_type == "Government", !is.na(prev2years_bingedrinking))


poissonmodel_govt <- pglm(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending  + 
                                 prev2years_bingedrinking + prev2years_druguse ,
                              #when controlling for all of the demand variables, nothing is significant. I think this is because sud and needing_tx are correlated with other variables (see above)
                              # +lag(sud)  + lag(needing_tx), 
    #the effect should be twooways because there may be state-level differences and country-level differences in each year                      
    effect = "twoways", 
   index = c('State_Name', 'year'),
    data = state_govt, 
    model = "within", 
   #model = 'within' means fixed effects, or variable intercept by state by constant slope
    #method = 'nr',
    family = poisson) #can consider using 'negbin' instead

summary(poissonmodel_govt)


simplemodel_govt <- pglm(count_tx ~ pop_thousands + medicaid_enroll_percapita + percapita_health_spending + lag(sud_pastyear), #when we include health spending all other significance goes away. why?
    effect = "individual",
    index = c('State_Name', 'year'),
    model = "within",
    data = state_govt, 
    family = poisson)

summary(simplemodel_govt)
exp(simplemodel_govt$estimate)
```

The population data from the CMS website (above) is the same as the FREDS data, but rounded to the nearest thousand 
```{r population data cleaning from FREDS data}

#statepop_wide <- read.table("Data/FRED_StatePopulations/State_Treatment_Analysis_Annual.txt", header = TRUE)

#change the data to have a row per state per year 
#statepop_long <- statepop_wide %>% 
# pivot_longer(cols = names(statepop_wide)[-1]) %>% 
#  mutate(state = gsub("POP", "", name)) %>%
#  rename(pop_thousands = value) %>%
##  separate(DATE, into = c("year", NA, NA), sep = "-") %>%
#  select(year, state, pop_thousands)

#statepop_long
```


