---
title: "SAMHSA Sensitivity Analysis"
output: html_notebook
---


Conduct sensitivity analysis to confirm whether these results hold under different assumptions


Assumption #1 - do the (direction of) the results still hold when including the other covariates (Medicaid, healthcare spending) as predictors?

Yes - the coefficients are almost exactly the same
```{r}
glmer_forprofit_sensitivity <- glmer.nb(count_tx ~ 
                                 prev2years_drugusecalc + prev2years_aud +
                                pop_millions +  medicaid_enroll_percapita +  percapita_health_spending_thousands +
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_forprofit)

summary(glmer_forprofit_sensitivity)
```


Assumption #2 - what do the relationships look like when the outcome is ONLY drug use and alcohol use? (no control variables, no random effects)

almost exactly the same 
```{r}
glmer_forprofit_sensitivity2 <- glmer.nb(count_tx ~ 
                                 prev2years_drugusecalc + prev2years_aud+
                                #pop_millions +  medicaid_enroll_percapita +  percapita_health_spending_thousands +
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_forprofit)

summary(glmer_forprofit_sensitivity2)
```


Assumption #3 - what happens if we separate out the cocaine and heroin variables? 


AUD stays negative and significant, but neither drug use variable is still significant
```{r}
glmer_forprofit_sensitivity3 <- glmer.nb(count_tx ~ 
                                  prev2years_aud + prev2years_heroin + prev2years_cocaine +
                                pop_millions + # medicaid_enroll_percapita +  percapita_health_spending_thousands +
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_forprofit)

summary(glmer_forprofit_sensitivity3)
```

Assumption #4 - what if we also include marijuana

Same general direction, but the drug use measurement increases to be slightly over 0.05
```{r}
glmer_forprofit_sensitivity4 <- glmer.nb(count_tx ~ 
                                 prev2years_drugusecalc + prev2years_aud + prev2years_marijuana +
                                pop_millions +  #medicaid_enroll_percapita +  percapita_health_spending_thousands +
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_forprofit)

summary(glmer_forprofit_sensitivity4)
```

Assumption #5 - what happens if we fit the model using the predictors that were best from LRT for nonprofit / govt?

Still not significant - this is a good thing because that was the result when we included AUD also 
```{r}
glmer_nonprofit_sensitivity1 <- glmer.nb(count_tx ~ 
                                 prev2years_drugusecalc +
                                pop_millions +  
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_nonprofit)
summary(glmer_nonprofit_sensitivity1)

glmer_govt_sensitivity1 <- glmer.nb(count_tx ~ 
                                 prev2years_drugusecalc +
                                pop_millions +  
                                (1|State_Name)+ (1 | year),
                              control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                              data = state_govt)
summary(glmer_govt_sensitivity1)
```




Assumption #6 - do these trends hold PRE- ACA? 
```{r}

```




Assumption # 7 - do results change if we include the (lagged) overdose rates? i.e. can we measure whether the number of treatment centers is a function of how much attention is given to drug use (brought by overdose deaths) and not by prevalence of substance use? 


First, look how closely drug use and overdose rates are correlated
```{r}
state_final_df %>%
  filter(is.na(overdose_rate_lag))
```

```{r}
cor(state_final_df[, c("pop_millions", "percapita_health_spending_thousands", "medicaid_enroll_percapita", "prev2years_drugusecalc", "prev2years_aud", "prev2years_marijuana", "Gender_Male", "Ethnicity_HispanicorLatino", "Race_White", "Insurance_Uninsured", "Unemployment", "Income", "overdose_rate_lag")], use = "complete.obs")
```

```{r}
overdose_controls_str <- "~ prev2years_aud + prev2years_drugusecalc + prev2years_marijuana +
                                overdose_rate_lag + 
                                scale(pop_millions) + 
                                scale(Income) +
                                scale(medicaid_enroll_percapita) +
                                scale(Gender_Male) +
                                scale(Unemployment) + 
                                scale(percapita_health_spending_thousands) +
                                scale(Insurance_Uninsured) + 
                                scale(Ethnicity_HispanicorLatino) + 
                                scale(Race_White) +
                                (1| State_Name) + (1 | year)"



#the results are the same for for-profit - except that AUD becomes no longer significant (barely)

#overdose rates themselves are not significant 

#drug use prevalence is still closely correlated
glmer_overdose_forprofit <- glmer.nb(formula(paste("Forprofit", overdose_controls_str)), control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                            data = state_final_df)

summary(glmer_overdose_forprofit)
```

```{r}

#nonprofit results are the same - no relationship between drug use prevalence and number of treatment centers, nor with overdose deaths
glmer_overdose_nonprofit <- glmer.nb(formula(paste("Nonprofit", overdose_controls_str)), control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                            data = state_final_df)

summary(glmer_overdose_nonprofit)

#summary(allFit(glmer_overdose_nonprofit))
```

```{r}
#nonprofit results are the same - no relationship between drug use prevalence and number of treatment centers, nor with overdose deaths
glmer_overdose_govt <- glmer.nb(formula(paste("Government", overdose_controls_str)), control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e6)),
                            data = state_final_df)

summary(glmer_overdose_govt)
```

